Ornaments as indicators of social changes and cultural practices in northeastern Taiwan before and after European colonial period

### 2019-12-09 analysis/paper/paper.Rmd
Historical contexts about global and regional trade network relating to local indigenous societies

---
title: "Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeast Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeast Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in Taiwan, the northeast region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes ornaments excavated from Kiwulan (700-1900 AD), the largest Iron Age settlement on the Yilan plain in northeast Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

# Historical Background of complex exchange network during the late Iron Age and early historical period

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered the historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans.

In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

@@

Currently the ornaments found in northeast Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, a large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to the early historical period was involved in complex exchange networks both on a regional and global scale.

# Excavations at Kiwulan in northeast Taiwan

@@

The European colonial presence in northern Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrant in 1768 [@Chen1963; @Ke1993].

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornament types and materials during the European period. Northeast Taiwan, Yilan, was involved in complex trading networks both in regional scale with other indigenous groups and Chinese merchants, and global scale with the European including the Dutch and the Spanish. The origin of ornaments might be multiple places mostly in Southeast Asia and introduced into northeast Taiwan by Chinese merchants before 17th century. Later, the trade activities became more frequent and intense in the 17th century that might be related to the colonial involvement of the Europeans based on the similar styles of golden beads found in Luzon that was once the colony of the Spanish during this period. In addition to ornaments, stonewares at Kiwulan that were also found on a Spanish shipwreck and Southeast Asia in relation to the Dutch and the Spanish trading routes, and the tobacco use were also believed introduced to Kiwulan by the Europeans [@Wang2007; @Chen2007]. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period in the early 19th century. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. Although historical records indicate that trade ornaments were prestige goods in prehistoric northeast Taiwan, the information might be biased that needs to be compared with archaeological evidence.

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. This provides an interpretation that Yilan social organisation went from corporate mode before the European arrival, then transitioned to network mode during the European presence. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeast Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by colonial activities or involved in the colonial economy centred in the major European colonies [@Trabert2017]. Kiwulan in northeast Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the borders of direct European colonial occupation. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Ornaments were viewed as prestige goods in the local indigenous culture before European contact. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signalling wealth and privileged trading connections among the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.

>>>>

---
title: "Ornaments as indicators of social changes and cultural practices in northeastern Taiwan before and after European colonial period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in Taiwan, the northeastern region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes ornaments excavated from Kiwulan (700-1900 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

# Ornaments in complex exchange network during the late Iron Age and early historical period

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered the historical period since the early 17th century due to the colonial presence of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, Taiwan was incorporated into Qing dynasty and a large wave of Han Chinese migrated to Yilan during the late 18th century.

The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of indigenous people to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

@@

Currently the ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicates multiple origins more likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, a large amount of golden beads at Kiwulan might be introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Based on the archaeological evidence and historical records, Northeastern Taiwan was involved in the regional network with East Asia in the late Iron age that Chinese merchants traded metal materials, clothes, and beads with local indigenous people in Taiwan for local resources and later was incorporated into more international scale due to the arrival of the Europeans. The common finds of foreign stoneware jars from Southeast Asia in local sites were frequently found in European shipwrecks that indicates the direct or indirect interaction. Despite the origin of some ornaments were traced to China, it is likely the large amount of them found at the 17th century sites were traded into due to indirect economic activities stimulated by the Europeans or direct European colonial activities.

# Excavations at Kiwulan in northeastern Taiwan

@@

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrant in 1768 [@Chen1963; @Ke1993].

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornament types and materials during the European period. Northeastern Taiwan, Yilan, was involved in complex trading networks both in regional scale with other indigenous groups and Chinese merchants, and global scale with the European including the Dutch and the Spanish. The origin of ornaments might be multiple places mostly in Southeast Asia and introduced into northeastern Taiwan by Chinese merchants before 17th century. Later, the trade activities became more frequent and intense in the 17th century that might be related to the colonial involvement of the Europeans based on the similar styles of golden beads found in Luzon that was once the colony of the Spanish during this period. In addition to ornaments, stonewares at Kiwulan that were also found on a Spanish shipwreck and Southeast Asia in relation to the Dutch and the Spanish trading routes, and the tobacco use were also believed introduced to Kiwulan by the Europeans [@Wang2007; @Chen2007]. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period in the early 19th century. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. Although historical records indicate that trade ornaments were prestige goods in prehistoric northeastern Taiwan, the information might be biased that needs to be compared with archaeological evidence.

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. This provides an interpretation that Yilan social organisation went from corporate mode before the European arrival, then transitioned to network mode during the European presence. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by colonial activities or involved in the colonial economy centred in the major European colonies [@Trabert2017]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the borders of direct European colonial occupation. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Ornaments were viewed as prestige goods in the local indigenous culture before European contact. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signalling wealth and privileged trading connections among the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.
